Parsing Postfix log files

### 2007-11-18 doc/logparser.tex
Finished the glossary entries for Postfix components.

The \SMTPlong, originally defined in \RFC{} 821~\cite{RFC821} and updated
in \RFC{} 2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{} 760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time.}  A typical \SMTP{}
conversation resembles the following (the lines starting with a three digit
number are sent by the server, all other lines are sent by the client):

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's \regex{}, generate a \DFA{} from it, and
examine every \DFA{} to determine if any overlap.  The author has not
attempted to do this: such a project by itself would probably qualify for a
PhD, and may involve solving the Halting
Problem~\cite{Wikipedia-halting-problem} and circumventing the
Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line is unspecified, and
the first matching rule wins.  Overlapping rules are frequently a
requirement, allowing a more specific rule to match some lines and a more
general rule to match the majority, e.g.\ separating \SMTP{} delivery to a
specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches when parsing the previous log file.  Negative
priorities may be useful for catchall rules.

@@

% Postfix components
\glossary{name={bounce},description={
    XXX The definitive documentation is \url{http://www.postfix.org/bounce.8.html}.
}}

@@

\glossary{name={local},description={
    XXX The definitive documentation is \url{http://www.postfix.org/local.8.html}.
}}

@@

\glossary{name={postdrop},description={
    XXX The definitive documentation is \url{http://www.postfix.org/postdrop.1.html}.
}}

\glossary{name={postsuper},description={
    XXX The definitive documentation is available at \url{http://www.postfix.org/postsuper.1.html}.
}}

@@

\glossary{name={sendmail},description={
    Postfix provides a command that is compatible with the Sendmail
    (\url{http://www.sendmail.org/}) mail submission program
    that all Unix commands which send mail depend on.  The definitive
    documentation is \url{http://www.postfix.org/sendmail.1.html}.
}}

\glossary{name={smtp},description={
    XXX The definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
}}

@@

\glossary{name={Bayesian poisoning},description={
    Bayesian poisoning is the addition of innocuous or unrelated words to a
    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
    word Viagra would be firmly in the spam corpus for most people, but by
    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
    forward\/} and \textit{best business practices\/} to a mail received by
    a manager, the Bayesian spam filter might tip the balance from bad to
    good, because the non-spam words might outweigh the spam words.
}}

\glossary{name={$<>$},sort={<>},description={
    $<>$ is the sender address used for mail which should not be replied
    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
    in $<>$, so \textit{username\at{}domain\/} becomes
    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
    address.  All mail servers must accept mail sent from $<>$, or they are
    in violation of \SMTP{}.
}}

>>>>

The \SMTPlong, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time.}  A typical \SMTP{}
conversation resembles the following (the lines starting with a three digit
number are sent by the server, all other lines are sent by the client):

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's \regex{}, and determine if any overlap.  The
author has not attempted to do this: such a project by itself would
probably qualify for a PhD, and may involve solving the Halting
Problem~\cite{Wikipedia-halting-problem} and circumventing the
Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line is unspecified, and
the first matching rule wins.  Overlapping rules are frequently a
requirement, allowing a more specific rule to match some lines and a more
general rule to match the majority, e.g.\ separating \SMTP{} delivery to a
specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches when parsing the previous log file.  Negative
priorities may be useful for catchall rules.

@@

% Postfix components
\glossary{name={bounce},description={
    The bounce daemon is responsible for sending bounce notifications in
    Postfix versions later than 2.2.  The definitive documentation is
    \url{http://www.postfix.org/bounce.8.html}.
}}

@@

\glossary{name={local},description={
    Local is the Postfix component responsible for local delivery of mail
    (i.e.\ delivered on the server Postfix is running on), whether it be to
    a user's mailbox or a program such as a mailing list manager or
    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
    processing of a user's \texttt{.forward} file.  The definitive
    documentation is \url{http://www.postfix.org/local.8.html}.
}}

@@

\glossary{name={postdrop},description={
    Postdrop is used when submitting mail locally on the server: it creates
    a new mail in the queue and copies its input into the mail.  Subsequent
    delivery of the mail is the responsibility of other Postfix components.
    The definitive documentation is
    \url{http://www.postfix.org/postdrop.1.html}.
}}

\glossary{name={postsuper},description={
    Maintenance task such as deleting mails from the queue, putting mail on
    hold (no further delivery attempts will be made until it is released
    from hold, also by postsuper), and consistency checking of the mail
    queue.  The definitive documentation is available at
    \url{http://www.postfix.org/postsuper.1.html}.
}}

@@

\glossary{name={sendmail},description={
    Postfix provides a command that is compatible with the Sendmail
    (\url{http://www.sendmail.org/}) mail submission program that all Unix
    commands which send mail depend on; Postfix sendmail executes postdrop
    to place a new mail in the queue.  The definitive documentation is
    \url{http://www.postfix.org/sendmail.1.html}.
}}

\glossary{name={smtp},description={
    Delivery of mail over \SMTP{} is performed by the smtp component.  The
    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
}}

@@

\glossary{name={Bayesian poisoning},description={
    Bayesian poisoning is the addition of innocuous or unrelated words to a
    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
    word Viagra would be firmly in the spam corpus for most people, but by
    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
    forward\/} and \textit{best business practices\/} to a mail received by
    a manager, the Bayesian spam filter might tip the balance from bad to
    good, if the non-spam words outweigh the spam words.
}}

\glossary{name={$<>$},sort={<>},description={
    $<>$ is the sender address used for mail which should not be replied
    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
    in $<>$, so \textit{username\at{}domain\/} becomes
    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
    address, but is always written as $<>$ for clarity.  All mail servers
    must accept mail sent from $<>$, or they are in violation of
    \RFC{}~2821~\cite{RFC2821}.
}}
